---
title: "Time Series"
author: "Emerson Amirhosein  Azarbakht"
output: pdf_document
---

# Time Series (TS)

#### Used in
- control of inventory, based on demand trends
- airline's decision to buy airplanes bc of passenger trends and decision to increase/maintain market share
- climate change decisions based on temperature change trends
- business/sales forecasting
- everyday operational decisions
- long-term effects of proposed water management policies by simulating daily rainfall and sea state time series
- understanding fluctuations in monthly sales
- basis for signal processing in telecommunications <?>
- disease incidence tracking, yearly rates
- census analysis
- tracking monthly unemployment rate; as an economic indicator used by decision makers

#### Used to
- to understand the past, and predict the future
- forcasting (predicting inference, a subset of statistical inference). assumes that present trends continue. This assumption cannot be checked empirically, but, when we identify the likely causes for a trend, we can justify the forecasting(extrapolating it) for a few time-steps at least
- anomaly detection
- clustering
- classification (assigning a time series pattern to a specific category: e.g. gesture recognition of hand movements in sign language videos)
- query by content ~ Content-based image retrieval

#### Data: 
a variable measured sequentially in time, or at a fixed [sampling] interval 

#### serial dependence problem:
observations close together in time tend to be correlated (serially dependent)

TS tries to explain this correlation (serial dependence)
autocorrelation analysis examines this serial dependence <?>

#### conditions (assumptions of TS)
- stationary process ?
- Ergodic process ?

```{r}
plot(AirPassengers)
start(AirPassengers)
end(AirPassengers)
frequency(AirPassengers)
plot(AirPassengers)
summary(AirPassengers)
layout(1:2)
# takes an input matrix for the location of each plot in the graphics window
plot(aggregate(AirPassengers))
boxplot(AirPassengers ~ cycle(AirPassengers))
```

plotting shows _patterns_, and _features_ of the data + *outliers* and *erroneous* values 


#### patterns
1. trend = a non-periodic systematic change in a TS
    * can be modeled simply by a linear increase or decrease. (only if it's deterministic ~ it's non-stochastic)
    * stochastic trend: seems to change direction at unpredictable times rather than displaying a consistent pattern (e.g. like the air passenger series)
2. seasonal variation = a repeating pattern within a fixed period (e.g. each year)
3. cycles = a non-fixed-period cycle (without a fixed period). example: El-Nino

```{r}
# monthly unemployment rate for the US state of Maine from January 1996 until August 2006
Maine.month <- read.table("http://staff.elena.aut.ac.nz/Paul-Cowpertwait/ts/Maine.dat", header = TRUE)
# header TRUE means treat first row as column names
attach(Maine.month)
str(Maine.month)
head(Maine.month)
class(Maine.month)
# it's a data.frame, not a ts object. So, we need to convert it to ts
Maine.month.ts <- ts(unemploy, start = c(1996, 1), freq = 12)
Maine.month.ts
Maine.annual.ts <- aggregate(Maine.month.ts)/12
layout(1:2)
plot(Maine.month.ts)
plot(Maine.annual.ts)
Maine.Feb <- window(Maine.month.ts, start = c(1996,2), freq = TRUE)
Maine.Aug <- window(Maine.month.ts, start = c(1996,8), freq = TRUE)
Feb.ratio <- mean(Maine.Feb) / mean(Maine.month.ts)
Aug.ratio <- mean(Maine.Aug) / mean(Maine.month.ts)
```

========

##### natoionwide unemployment rate
```{r}
# monthly unemployment rate for all of the United States from January 1996 until October 2006
US.month <- read.table("http://staff.elena.aut.ac.nz/Paul-Cowpertwait/ts/USunemp.dat", header = T)
attach(US.month)
US.month.ts <- ts(USun, start=c(1996,1), end=c(2006,10), freq = 12)
plot(US.month.ts, ylab = "unemployed (%)")
```
========

#### Multiple TS 
```{r}
ChocolateBeerElectricity <- read.table("http://staff.elena.aut.ac.nz/Paul-Cowpertwait/ts/cbe.dat", header = TRUE)
class(ChocolateBeerElectricity)
str(ChocolateBeerElectricity)
Chocolate.ts <- ts(ChocolateBeerElectricity[,1], start = 1958, frequency = 12)
Beer.ts <- ts(ChocolateBeerElectricity[,2], start = 1958, frequency = 12)
Electricity.ts <- ts(ChocolateBeerElectricity[,3], start = 1958, frequency = 12)

plot(cbind(Chocolate.ts, Beer.ts, Electricity.ts))

```

========
#### Climate change
```{r}
# Global temperature anomalies from the monthly means over the period
GlobalTemperatures <- scan("http://staff.elena.aut.ac.nz/Paul-Cowpertwait/ts/global.dat")
# scan to read data into a vector or list from the console or file.
?scan
str(GlobalTemperatures)
GlobalTemperatures.ts <- ts(GlobalTemperatures, st = c(1856,1), end = c(2005,1), frequency = 12)
head(GlobalTemperatures.ts)
tail(GlobalTemperatures.ts)
Global.annual <- aggregate(GlobalTemperatures.ts, FUN = mean)
head(Global.annual)
Global.annual
plot(GlobalTemperatures.ts)
plot(Global.annual)  

#
New.series <- window(GlobalTemperatures.ts, start=c(1970, 1), end=c(2004, 12))
New.time <- time(New.series)
plot(New.series)
abline(reg = lm(New.series ~ New.time))
```

